Persistence in functional topology and a correction to a theorem of Morse

### 2021-07-16 inequalities.tex
Changes during meeting

In this section we prove that $q$-tame persistence modules satisfy a general version of Morse inequalities that specializes to the usual Morse inequalities in the smooth context, as well as to the version used by Morse--Tompkins to prove their Unstable Minimal Surface Theorem, see \cref{s:surfaces} for more details on this result.
We deduce these general inequalities from the existence of persistence diagrams as reviewed in \cref{thm:q-tame modules have barcodes}.

Recall that for a Morse function $f$ on a closed manifold $X$, the usual Morse inequalities state that for any $n$ we have
\begin{equation}\label{equ:morse_inequalities}
    \sum_{d=0}^n \ (-1)^{n-d} ( \# \crit_{d}(f) - \beta_{d}(X) ) \ \geq\  0,  
\end{equation}
where $\crit_{d}(f)$ is the set of critical points of $f$ with index $d$ and $\beta_{d}(X)$ is the $d$th Betti number of $X$.

Now if no two critical points of $f$ have the same value, the critical points are in one-to-one correspondence with the critical values, which in turn are in one-to-one correspondence with the homological changes in the sublevel set filtration of $f$, i.e., the endpoints of the intervals appearing the barcode of $H_*(f_{\leq \bullet})$.

More precisely, an index $d$ critical point may either kill an existing homology class, in which case it corresponds to the right endpoint of an interval in the barcode of $H_{d-1}(f_{\leq \bullet})$, or it may give rise to a new homology class, in which case it corresponds to the left endpoint of an interval in the barcode of $H_d(f_{\leq \bullet})$.

The Betti numbers of $X$ may also be expressed in terms of barcodes because they agree with the number of intervals that extend to $+\infty$. 
Thus, the Morse inequalities above can be expressed entirely in terms of the barcode or persistence diagram of the sublevel set filtration of the function, which encodes the homological changes in the filtration.

This approach of counting homological changes instead of critical points is also what Morse used in his non-smooth setting of functional topology.
To keep track of the number of $d$-dimensional homological events at filtration value $t$ that persist for at least time $e$ but not indefinitely, \citet{Morse.1940} defined the $(d, t, e)$-\textit{cap numbers} of a filtration, which, when the associated persistence module has a barcode and is upper semi-continuous\footnote{A persistence module $M$ is \emph{upper semi-continuous} if $M_{t} \to \lim_{u > t} M_{u}$ is an isomorphism for all~$t$.}, correspond to the number of finite bars with length greater than $e$ in the $d$-dimensional barcode with left endpoint $t$, plus the number of finite bars with length greater than $e$ in the $(d-1)$-dimensional barcode with right endpoint $t$ (killed by something $d$-dimensional).

Working entirely in the algebraic setting, we will now fix a $\Z$-graded q-tame persistence module $M$.
Of course, one may think of $M$ as the persistent homology of a q-tame filtration, but $M$ could for example also arise as the filtered Floer homology of some Hamiltonian on a symplectic manifold.

By \cref{thm:q-tame modules have barcodes} $M$ has a persistence diagram in every degree $d$ with multiplicity function denoted $\mathfrak{m}_d \colon \multiplicityDomain \to \N$.
In analogy to Morse's definitions, we may define the $(d, t, e)$-\textit{cap number} of our graded q-tame persistence module $M$ in terms of its persistence diagram as
\[
m_{d}^{> e}(t) =
\sum_{\substack{p \in \R \\ t - p > e}} \mathfrak{m}_{d-1}(p, t) +
\sum_{\substack{q \in \R \\ q - t > e}} \mathfrak{m}_d(t, q).
\]
Note that finiteness of these quantities is ensured by the q-tameness of $M$. 
Moreover, note that the condition on the counted features to have persistence greater than $e$ is necessary for finiteness of the cap numbers in the q-tame setting. 
One may omit the lower persistence bound $e$ in the definition if one restricts to $\PFD$ persistence modules.

>>>>

In this section we prove that $q$-tame persistence modules satisfy a general version of Morse inequalities that specializes to the usual Morse inequalities in the smooth context, as well as to the version used by Morse--Tompkins in their proof of the Unstable Minimal Surface Theorem (see \cref{s:surfaces} for more details on this result).
We deduce these general inequalities from the existence of persistence diagrams as reviewed in \cref{thm:q-tame modules have barcodes}.

Recall that the usual Morse inequalities state that for a Morse function $f$ on a closed manifold $X$ and for any $n \in \mathbb N$ we have
\begin{equation}\label{equ:morse_inequalities}
    \sum_{d=0}^n \ (-1)^{n-d} ( \# \crit_{d}(f) - \beta_{d}(X) ) \ \geq\  0,  
\end{equation}
where $\crit_{d}(f)$ is the set of critical points of $f$ with index $d$ and $\beta_{d}(X)$ is the $d$th Betti number of $X$.

Now if $f$ is a \emph{excellent} Morse function, meaning that no two critical points of $f$ have the same value, the critical points are in one-to-one correspondence with the critical values, which in turn are in one-to-one correspondence with the homological changes in the sublevel set filtration of $f$, i.e., the endpoints of the intervals appearing the barcode of $H_*(f_{\leq \bullet})$. 
More precisely, an index $d$ critical point may either kill an existing homology class in degree $d-1$, in which case it corresponds to the right endpoint of an interval in the barcode of $H_{d-1}(f_{\leq \bullet})$, or it may give rise to a new homology class in degree $d$, in which case it corresponds to the left endpoint of an interval in the barcode of $H_d(f_{\leq \bullet})$.
The Betti numbers of $X$ may also be expressed in terms of barcodes, as they agree with the number of intervals that extend to $+\infty$. 
Thus, the Morse inequalities above can be expressed entirely in terms of the barcode or persistence diagram of the sublevel set filtration of the function, which encodes the homological changes in the filtration.

This approach of counting homological changes instead of critical points is also what Morse used in his non-smooth setting of functional topology.
To keep track of the number of $d$-dimensional homological events at filtration value $t$ that persist for at least time $e$ but not indefinitely, \citet{Morse.1940} defined the $(d, t, e)$-\textit{cap numbers} of a filtration, which can be described in terms of persistent homology as follows.
Assuming that the associated persistence module $M$ has a barcode and is \emph{upper semi-continuous}, meaning that $M_{t} \to \lim_{u > t} M_{u}$ is an isomorphism for all~$t$, the $(d, t, e)$-cap number correspond to the number of finite bars with length greater than $e$ in the $d$-dimensional barcode with left endpoint $t$, plus the number of finite bars with length greater than $e$ in the $(d-1)$-dimensional barcode with right endpoint $t$ (killed by something $d$-dimensional).

Working entirely in the algebraic setting, we will now fix a $\Z$-graded q-tame persistence module $M$.
Of course, one may think of $M$ as the persistent homology of a q-tame filtration, but $M$ could also arise, for example, as the filtered Floer homology of some Hamiltonian on a symplectic manifold.

By \cref{thm:q-tame modules have barcodes}, the persistence module $M$ has a persistence diagram in every degree $d$, with multiplicity function denoted $\mathfrak{m}_d \colon \multiplicityDomain \to \N$.
In analogy to Morse's definitions, we may define the $(d, t, e)$-\textit{cap number} of our graded q-tame persistence module $M$ in terms of its persistence diagram as
\[
m_{d}^{> e}(t) =
\sum_{\substack{p \in \R \\ t - p > e}} \mathfrak{m}_{d-1}(p, t) +
\sum_{\substack{q \in \R \\ q - t > e}} \mathfrak{m}_d(t, q).
\]
Note that finiteness of these quantities is ensured by the q-tameness of $M$. 
Moreover, note that the condition on the counted features to have persistence greater than $e$ is necessary for finiteness of the cap numbers in the q-tame setting. 
One may omit the lower persistence bound $e$ in the definition if one restricts to $\PFD$ persistence modules.

### 2021-07-16 persistence.tex
Improve sect 4
Changes during meeting

In this section we present an overview of the theory of persistence as it is used in geometric contexts, through filtrations by sublevel sets of real-valued functions.
For a detailed exposition we refer to \cite{polterovich2020topological} and \cite{Chazal.2016a, MR3408277}.

@@

Unfortunately, the $\PFD$ condition is too restrictive for many purposes.
In particular, it is unsuited for the the applications of Morse and Tompkins in minimal surfaces theory.
An appropriate weakening of $\PFD$ for more general settings is the notion of \emph{q-tame}.
A persistence module $M$ is q-tame if the rank of the map $M_{s,t} \colon M_s \to M_t$ is finite dimensional for all $s < t$  \cite{Chazal.2016a}.
As exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$, not every q-tame persistence module admits a barcode decomposition in the above sense.
Yet, there are multiple ways of working around this to still get discrete invariants in the spirit of barcodes \cite{Chazal.2016a, Chazal.2016b, schmahl2020structure}.
We briefly recall the approach from \citet{Chazal.2016b}.

A persistence module $K$ is called \emph{ephemeral} if the maps $K_{s,t} \colon K_s \to K_t$ are $0$ for all $s < t$.
The \emph{radical} $\rad M$ of a persistence module $M$ is defined as the unique minimal submodule of $M$ such that the cokernel of the inclusion $\rad M \to M$ is an ephemeral persistence module.
Specifically, $(\rad M)_t = \sum_{s<t}\im M_{s,t}$.
If $M$ is q-tame, then its radical admits a barcode decomposition \cite[Corollary~3.6]{Chazal.2016b},
with the associated barcode describing the isomorphism type of~$M$ ``up to ephemerals".
This can be formalized by constructing the \emph{observable category of persistence modules}, which is equivalent to the quotient of the category of persistence modules by the full sub-category of ephemeral persistence modules.
Intuitively, one may think of the observable category as forgetting all information in persistence modules that does not persist over a non-zero amount of time.
The barcode of the radical of a q-tame persistence module $M$ is then a complete invariant of $M$ in the observable category.

Instead of talking about the barcode of the radical of a q-tame persistence module, it will be more convenient for our treatment of Morse inequalities to talk about the \emph{(undecorated) persistence diagram} associated to a barcode $\{I_{\lambda}\}_{\lambda \in \Lambda}$, which is the multiset defined by the \emph{multiplicity function} $\mathfrak{m} \colon \multiplicityDomain \to \mathbb{N}$ that associates to an element in
\[
\multiplicityDomain =
\big\{ (p,q) \mid p \in \R \cup \{-\infty\}, \ q \in \R \cup \{+\infty\}, \ p < q \big\}
\]
the cardinality of the set $\{ \alpha \in A \mid \inf I_{\alpha} = p,\ \sup I_{\alpha} = q\}$.

@@

The most general stability result is shown by considering as an intermediate step the metric space of filtrations and persistence modules equipped with the \emph{interleaving distance} \cite{MR2279866}.
Given two functors $M,N \colon \R \to \mathsf C$, where $\mathsf C$ is typically $\Top$ or $\Vect$, a $\delta$-interleaving consists of a pair of natural transformations $(M_t \to N_{t+\delta})_t$ and $(N_t \to M_{t+\delta})_t$ from one diagram to a shifted version of the other and vice versa, which compose to the structure maps $(M_t \to M_{t+2\delta})_t$ and $(N_t \to N_{t+2\delta})_t$.
Clearly, the case $\delta=0$ describes an isomorphism, and the infimum $\delta$ for which $M$ and $N$ admit a $\delta$-interleaving is defined as the interleaving distance between them.
As it turns out, the stability of persistence barcodes can then be described as a sequence of 1-Lipschitz transformations: from functions (with the supremum norm) to filtrations by sublevel sets (with the interleaving distance on diagrams of topological spaces), to persistent homology (with interleaving distance on persistence modules), and finally to persistence diagrams (with the bottleneck distance).
In this approach, by far the most difficult step is showing that passing from persistence modules to persistence diagrams is stable, a result which is known as the Algebraic Stability Theorem  \cite{10.1145/1542362.1542407,Chazal.2016a,MR3333456}.

>>>>

In this section we present an overview of the theory of persistence through filtrations by sublevel sets of real-valued functions, applicable to functionals defined on a function or moduli space.
For a detailed exposition we refer to \cite{polterovich2020topological} and \cite{Chazal.2016a, MR3408277}.

@@

Unfortunately, the $\PFD$ condition is too restrictive for many purposes.
In particular, it is unsuited for the the applications of Morse and Tompkins in minimal surfaces theory.
An appropriate weakening of $\PFD$ for more general settings is the notion of \emph{q-tame}.
A persistence module $M$ is q-tame if the rank of the map $M_{s,t} \colon M_s \to M_t$ is finite for all $s < t$ \cite{Chazal.2016a}.
As exemplified by the infinite product of interval modules $\prod_{n \in \N_{> 0}} C([0,1/n))$, not every q-tame persistence module admits a barcode decomposition in the above sense.
Yet, there are multiple ways of working around this to still get discrete invariants in the spirit of barcodes \cite{Chazal.2016a, Chazal.2016b, schmahl2020structure}.
We briefly recall the approach from \citet{Chazal.2016b}.

A persistence module $K$ is called \emph{ephemeral} if the maps $K_{s,t} \colon K_s \to K_t$ are $0$ for all $s < t$.
The \emph{radical} $\rad M$ of a persistence module $M$ is the unique minimal submodule of $M$ such that the cokernel of the inclusion $\rad M \to M$ is an ephemeral persistence module.
Specifically, $(\rad M)_t = \sum_{s<t}\im M_{s,t}$.
If $M$ is q-tame, then its radical admits a barcode decomposition \cite[Corollary~3.6]{Chazal.2016b},
with the associated barcode describing the isomorphism type of~$M$ ``up to ephemerals".
This can be formalized by constructing the \emph{observable category of persistence modules}, which is equivalent to the quotient of the category of persistence modules by the full sub-category of ephemeral persistence modules.
Intuitively, one may think of the observable category as forgetting all information in persistence modules that does not persist over a non-zero amount of time.
The barcode of the radical of a q-tame persistence module $M$ is then a complete invariant of $M$ in the observable category.

Instead of talking about the barcode of the radical of a q-tame persistence module, it will be more convenient for our treatment of Morse inequalities to talk about the \emph{(undecorated) persistence diagram} associated to a barcode $\{I_{\lambda}\}_{\lambda \in \Lambda}$, which is the multiset defined by the \emph{multiplicity function} $\mathfrak{m} \colon \multiplicityDomain \to \mathbb{N}$ that associates to an element in the set $\multiplicityDomain$ of pairs of interval endpoints,
\[
\multiplicityDomain =
\big\{ (p,q) \mid p \in \R \cup \{-\infty\}, \ q \in \R \cup \{+\infty\}, \ p < q \big\},
\]
the cardinality of the set $\{ \alpha \in A \mid \inf I_{\alpha} = p,\ \sup I_{\alpha} = q\}$.

@@

The most general stability result is shown by considering as an intermediate step the metric spaces of filtrations and persistence modules equipped with the \emph{interleaving distance} \cite{MR2279866}.
Given two functors $M,N \colon \R \to \mathsf C$, where $\mathsf C$ is typically $\Top$ or $\Vect$, a $\delta$-interleaving consists of a pair of natural transformations $(M_t \to N_{t+\delta})_t$ and $(N_t \to M_{t+\delta})_t$ from one diagram to a shifted version of the other and vice versa, which compose to the structure maps $(M_t \to M_{t+2\delta})_t$ and $(N_t \to N_{t+2\delta})_t$.
Clearly, the case $\delta=0$ describes an isomorphism, and the infimum $\delta$ for which $M$ and $N$ admit a $\delta$-interleaving is defined as the \emph{interleaving distance} between them.
As it turns out, the stability of persistence barcodes can then be described as a sequence of several 1-Lipschitz transformations: from functions (with the supremum norm) to filtrations by sublevel sets (with the interleaving distance on diagrams of topological spaces), to persistent homology (with interleaving distance on persistence modules), and finally to persistence diagrams (with the bottleneck distance).
In this approach, by far the most difficult step is showing that passing from persistence modules to persistence diagrams is stable, a result which is known as the Algebraic Stability Theorem  \cite{10.1145/1542362.1542407,Chazal.2016a,MR3333456}.
